One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-10-01 papers/1_dmatpol/revision/paper-revision-1.tex
Revision files for DMS paper added

%%
%%   Version 3.1 of 31 July 2018.
%%
\documentclass[aip,jcp,preprint,amsmath,amssymb,floatfix]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

@@

However, even with the modern advances in computational chemistry, accurate description of phenomena due 
to complex heterogeneous molecular environments remains a challenge in the following areas: (i) describing 
the electronically excited states (electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017,
Xu.Blasiak.Cho.Layfield.Londergan.JPCL.2018,Lewis.Gruenke.Oliver.Ballottari.Bassi.Fleming.JPCL.2016} 
Interface between QM and MM levels is usually realized by electrostatic embedding and only rarely 
the Pauli exclusion principle and dispersion are additionally included.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
due to complicated nature of intermolecular interactions that theoretically cannot be solely envisioned 
as a sum of separate contributions coming from electrostatic, dispersion and Pauli 
effects.\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
For example,
one of the most frequently used, efficient and reasonably accurate \emph{ab initio} fragmentation methods, 
the effective fragment potential (EFP) methods, are limited to the Hartree\hyp{}Fock 
(HF)\cite{Roothaan.RevModPhys.1951,Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Hohenberg.Kohn.PhysRev.1964,Kohn.Sham.PhysRev.1965,Nguyen.Pachter.Day.JCP.2014}.
Unfortunately, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.

@@

Polarization under the uniform electric field is the simplest type of electrostatic perturbation
whereby it is possible to compute the exact DMS according to Eq.~\eqref{e:dD-Taylor} 
directly by numerical differentiation. Here, the DMS's up to second order are considered, i.e.,
%
\begin{subequations}\label{e:exact-DMS}
 \begin{align}
  {\bf B}^{(1)}_{\alpha\beta} &= \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B10} \\
  {\bf B}^{(2)}_{\alpha\beta} &= \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B20}
 \end{align}
\end{subequations}
%
and electric field magnitudes in a range from very weak ($<$0.001 a.u.) to very strong ($\sim$0.07 a.u.).
The \emph{ab initio} model from Eq.~\eqref{e:final-model.HF} and
the generalized model from Eq.~\eqref{e:final-model.General} are also discussed for the sake of comparison. 
The associated error distributions in prediction of polarization energies 
are shown in scatter plot in Figure~\ref{f:fig-1}, whereas the
statistical performance descriptors of polarization energies and induced multipole moments
are gathered in Table~\ref{t:results}.
%
\begin{figure}[t]
\includegraphics[width=0.5\textwidth]{fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the DMS tensors of the ground\hyp{}state water molecule
in uniform electric field.} 
$R_{\rm max}$ denotes the maximum rank of the electric field
in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0.0 to 0.07 a.u. were tested in 100 samples with randomly selected 
directions of electric field.
} 
\end{figure}
%

@@

>>>>

%%
%%   Version 3.1 of 31 July 2018.
%%
\documentclass[aip,jcp,reprint,amsmath,amssymb,floatfix]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

@@

However, even with the modern advances in computational chemistry, accurate description of phenomena due 
to complex heterogeneous molecular environments remains a challenge in the following areas: (i) describing 
the electronically excited states (electronic solvatochromism),\cite{Barbati.JACS.2014,
Szabla.Sponer.Jiri.Gora.JPCL.2015,Bednarska.Zalesny.Tian.Murugan.Agren.Bartkowiak.Molecules.2017,
Jedrzejewska.Grabarz.Bartkowiak.Osmialowski.SpectChimActA.2018} 
(ii) understanding the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms).\cite{Blasiak.Londergan.Webb.Cho.ACR.2017,
Xu.Blasiak.Cho.Layfield.Londergan.JPCL.2018,Lewis.Gruenke.Oliver.Ballottari.Bassi.Fleming.JPCL.2016} 
Interface between QM and MM levels is usually realized by electrostatic embedding and only rarely 
the Pauli exclusion principle and dispersion are additionally included.\cite{List.Olsen.Kongsted.PCCP.2016}
Unfortunately, it is quite difficult to efficiently treat the environment\hyp{}induced potential operator
due to complicated nature of intermolecular interactions that theoretically cannot be solely envisioned 
as a sum of separate contributions coming from electrostatic, dispersion and Pauli 
effects.\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994} 
For example,
one of the most frequently used, efficient and reasonably accurate \emph{ab initio} fragmentation methods, 
the effective fragment potential (EFP) methods, are limited to the Hartree\hyp{}Fock 
(HF)\cite{Roothaan.RevModPhys.1951,Gordon.Smith.Xu.Slipchenko.AnnuRevPhysChem.2013} 
or density functional theory (DFT) levels\cite{Hohenberg.Kohn.PhysRev.1964,Kohn.Sham.PhysRev.1965,Nguyen.Pachter.Day.JCP.2014}.
While they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states. 
In such cases, the full QM method needs to be utilized on the excited state wavefunction, that is 
embedded within the EFP environment through the electrostatic or polarizable embedding.\cite{Ghosh.JPCA.2017}

@@

Polarization under the uniform electric field is the simplest type of electrostatic perturbation
whereby it is possible to compute the exact DMS according to Eq.~\eqref{e:dD-Taylor} 
by using either the coupled\hyp{}perturbed Hartree\hyp{}Fock method\cite{Horn.Head-Gordon.JCP.2015}
or numerical differentiation. Here, the DMS's up to second order are considered, i.e.,
%
\begin{subequations}\label{e:exact-DMS}
 \begin{align}
  {\bf B}^{(1)}_{\alpha\beta} &= \frac{\partial D_{\alpha\beta}}{\partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B10} \\
  {\bf B}^{(2)}_{\alpha\beta} &= \frac{1}{2} 
     \frac{\partial^2 D_{\alpha\beta}}{\partial {\bf F} \otimes \partial {\bf F}}\Bigg|_{{\bf F}={\bf 0}} \;, \label{e:B20}
 \end{align}
\end{subequations}
%
and electric field magnitudes in a range from very weak ($<$0.001 a.u.) to very strong ($\sim$0.07 a.u.).
The \emph{ab initio} model from Eq.~\eqref{e:final-model.HF} and
the generalized model from Eq.~\eqref{e:final-model.General} are also discussed for the sake of comparison. 
The associated error distributions in prediction of polarization energies 
are shown in scatter plot in Figure~\ref{f:fig-1}, whereas the
statistical performance descriptors of polarization energies and induced multipole moments
are gathered in Table~\ref{t:results}.
%
\begin{figure}[t]
\includegraphics[width=0.5\textwidth]{fig-1.eps}
\caption{\label{f:fig-1} {\bf Performance of the DMS tensors of the ground\hyp{}state water molecule
in uniform electric field.} 
$R_{\rm max}$ denotes the maximum rank of the electric field
in the expansion from Eq.~\eqref{e:final-model.General}.
Electric field magnitudes in a range from 0.0 to 0.07 a.u. were tested in 100 samples with randomly selected 
directions of electric field.
} 
\end{figure}
%

@@

\subsection{\label{s:computational-cost}Computational cost and implementation of the DMS models}

Although the cost of computation of the perturbed one\hyp{}electron density matrices
from the DMS tensors and the distribution of the electric field is negligible, 
obtaining the generalized DMS matrix elements for a particular molecule
involves substantial amount of full QM calculations necessary to fit the DMS tensor elements.
This is, however, a relatively expensive operation that has to be performed only once.
%To get statistically relevant set of perturbed states, it is recommended to perform at least 100 
%full QM calculations. 
Since each of such preparatory calculations is an independent computational task, 
its automatization and parallelization is straightforward. Moreover, the implementation
of the method does not depend on the level of theory because only the perturbed density matrices
are needed. The computation of the gradient 
and Hessian matrix is inexpensive and its efficiency is limited only by the speed of Hessian matrix inversion.
Therefore, the method can be applied to all molecular fragments at any level of theory, provided
the one\hyp{}particle density matrices are available. In the illustrations shown in this work,
a set of 100 independent HF/6-311++G** calculations was sufficient to get statistically relevant set of perturbed 
ground state density matrices. This task was concluded within approximately 5 minutes per each DMS fit
on a single\hyp{}CPU computer.
